---
title: 'Reproducible Research: Peer Assessment 1'
output: html_document
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data
```{r}
# Read data
activity <- read.csv("activity.csv")
# Remove NAs
activityClean <- activity[complete.cases(activity),]
```

## What is the mean total number of steps taken per day?
```{r,warning=FALSE,message=FALSE}
library(dplyr)
# Calculate the total number of steps taken per day
sumSteps <- aggregate(steps~date,activityClean,sum)
# Make a histogram of the total number of steps taken each day
histSteps <- hist(sumSteps$steps, main="Histogram of the Total Number of Steps", xlab="Steps per Day",col="blue")
# Calculate and report the mean and median of the total number of steps taken per day
library(scales)
meanClean <- mean(sumSteps$steps)
meanClean <- comma_format(digits=5)(meanClean)
medianClean <- median(sumSteps$steps)
medianClean <- comma(medianClean)
```
* The `mean` of the total number of steps taken per day is **`r meanClean`.**

* The `median` of the total number of steps taken per day is **`r medianClean`.**

## What is the average daily activity pattern?
```{r}
# Make a time series plot of the 5-minute interval 
meanSteps <- aggregate(steps~interval,activityClean,mean)
plot(meanSteps$interval,meanSteps$steps,type="l",main="Average Number of Steps per Interval",xlab="Interval",ylab="Average Steps",col="blue")
```

The interval that contains the maximum number of steps is:
```{r}
maxInterval <- meanSteps[which.max(meanSteps$steps),]
maxInterval
```

## Imputing missing values
```{r}
# Compute total number of missing values
TotalMissing <- sum(is.na(activity))
TotalMissing <- comma(TotalMissing)
```
* The total number of missing values is **`r TotalMissing`**.

* To fill in all missing values, the strategy is to replace them with the average steps calculated per interval as shown below.
```{r}
# Fill in all missing values based on the average steps per interval
for (i in 1:nrow(activity)) {
  if(is.na(activity$steps[i])) {
    value <- meanSteps$steps[which(meanSteps$interval == activity$interval[i])]
    activity$steps[i] <- value
  }
}
activity$steps <- round(activity$steps,digits=0)
```

* The histogram of the new dataset with imputed values is then compared to the histogram of the old dataset where the null values are removed.
```{r,warning=FALSE,message=FALSE}
# Recalculate the mean, median, and histogram with imputed values
sumStepsImputed <- aggregate(steps~date,activity,sum)

# Make a histogram of the total number of steps taken each day with imputed values
par(mfrow=c(2,1),mar=c(5,4,2,1)) 
histSteps <- hist(sumSteps$steps, main="Histogram of the Total Number of Steps before Imputing", xlab="Steps per Day",col="blue")
histStepsImputed <- hist(sumStepsImputed$steps, main="Histogram of the Total Number of Steps after Imputing",xlab="Steps per Day",col="blue")

# Calculate and report the mean and median of the total number of steps taken per day wuth imputed values
meanImputed <- mean(sumStepsImputed$steps)
meanImputed <- comma_format(digits=5)(meanImputed)
medianImputed <- median(sumStepsImputed$steps)
medianImputed <- comma(medianImputed)
```

* The `mean` of the new dataset is **`r meanImputed`** vs the old mean of **`r meanClean`**.

* The `median` of the new dataset is **`r medianImputed`** vs the old median of **`r medianClean`**.

* *It can be observed that in terms of the histogram, mean and median, there is no significant difference between the two datasets.* 

## Are there differences in activity patterns between weekdays and weekends?
```{r,warning=FALSE,message=FALSE}
#Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
library(chron)

for(i in 1:nrow(activity)){ ### Order type of day
  if(is.weekend(activity$date[i])=="TRUE"){ 
    activity$type_of_day[i]<-"Weekend"
  }
  else { 
    activity$type_of_day[i]<-"Weekday"
  }
}

#Make a panel plot containing a time series plot of the 5-minute interval averaged across all weekday days or weekend days (y-axis)
activity$type_of_day <- as.factor(activity$type_of_day)

meanStepsWeekend <- aggregate(steps~interval+type_of_day,activity,mean)
library(ggplot2)
g <- ggplot(meanStepsWeekend, aes(interval, steps))
g + geom_line(stat = "identity", aes(color = type_of_day)) + 
  facet_grid(type_of_day~.) +
  labs(x="Interval", y="Steps",title= "Number of Steps per Interval")
```


* It can be observed that the activity pattern during weekends is fluctuating more than weekdays.